K2 Analysis

### 2014-04-22 text/micha_stat_analysis/notes_new.tex

@@

\subfloat[][\as neutral]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 29 &  7 & 0 & 1 \\
    Pos 2 & 26 & 10 & 0 & 1 \\
    Pos 3 & 25 &  9 & 1 & 2 \\ \addlinespace[0.15cm]
    Total & 80 & 26 & 1 & 4
\end{tabular}}
%
\hspace{0.2cm}
%
\subfloat[][\es neutral]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 16 &  7 & 0 & 14 \\
    Pos 2 & 21 & 13 & 0 &  3 \\
    Pos 3 & 16 & 13 & 0 &  8 \\ \addlinespace[0.15cm]
    Total & 53 & 33 & 0 & 25
\end{tabular}}

\subfloat[][\as accented]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 28 &  7 & 1 & 1 \\
    Pos 2 & 25 &  8 & 1 & 3 \\
    Pos 3 & 24 & 10 & 0 & 3 \\ \addlinespace[0.15cm]
    Total & 77 & 25 & 2 & 7
\end{tabular}}
%
\hspace{0.2cm}
%
\subfloat[][\es accented]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 13 &  6 & 0 & 18 \\
    Pos 2 & 24 &  7 & 1 &  5\\
    Pos 3 & 19 & 13 & 0 &  5 \\ \addlinespace[0.15cm]
    Total & 56 & 26 & 1 & 28
\end{tabular}}

@@

Two things strike the eye. Firstly, there are hardly any \emph{global}
responses. For subsequent analysis in terms of log-linear models we
must therefore lump together \emph{global} and \emph{false} responses
into one more general category level \emph{other}. Secondly, the
number of \emph{false} responses drops remarkably in the
\es-conditions. The following analysis should probe into this
potential position effect.

To test which factors are relevant to explain the data distribution in
Table~\ref{tab:CC-ContingencyTable}, we compute log-linear models
\citep{KnokeBurke1980:Log-Linear-Mode}. The relevant factors are
\textsc{Position}, \textsc{Response}, \textsc{Condition}, and
\textsc{Accent}. To ensure applicability, we lump levels \emph{global}
and \emph{false} of factor \textsc{Response} into a single level
\emph{other}. The best predictor of the data takes all factors
\textsc{Position}, \textsc{Response} and \textsc{Condition} and all
their interactions, including the three-way interaction, into account
($\chi^2 = 7.493$, $df=18$, $p = 0.985$, $AIC = 43.493$ (compared to
$AIC=72$ of the saturated model)). Crucially, factor \textsc{Accent}
is spurious.

% To test whether the effect of \textsc{Condition} could be due to the
% relatively high number of \emph{false} responses in the
% \es-conditions, we regrouped the factor \textsc{Response} once more
% to distinguish only \emph{local} responses from \emph{other}
% responses. Doing so allows us to address the question whether the
% amount of \emph{local} responses differed across conditions and
% accentuation types. But this does not seem to be the case. The model
% that predicts the data best in this case contains factors
% \textsc{Response} and \textsc{Position} and their two-way
% interaction ($\chi^2 = 4.71$, $df=18$, $p = 0.999$, $AIC = 16.711$
% (compared to $AIC=48$ of the saturated model)). This suggests that a
% shift in the distribution of \emph{local} responses can be
% satisfactorily explained as a position effect that is constant
% across conditions: later trials yielded less \emph{other} responses.

To test, then, whether \textsc{Position} affects the distribution of
\emph{literal} and \emph{local} readings, we also computed the best
log-linear predictor for the subset of our data including only
\emph{literal} and \emph{local} responses. Doing so, the best
predictor indeed drops factor \textsc{Position} and contains only
factors \textsc{Response} and \textsc{Condition} and their two-way
interaction ($\chi^2 = 12.497$, $df=20$, $p = 0.898$, $AIC = 20.497$
(compared to $AIC=48$ of the saturated model)). This suggests that the
factor \textsc{Position} is not relevant for explaining the
distribution of \emph{literal} and \emph{local} readings. Its effects
noted previously are most likely due to the decrease of \emph{false}
responses in later trials of the \es-conditions.

\todo[inline]{is this convincing enough to dispense with any potential
doubt about interferences from our screwed up block design?}

@@

Some subjects saw three others four trials of preference-related
controls. The relevant factors in this case are: \textsc{Condition}
with levels \emph{lc} and \emph{ec} coding which reading could be
judged first in a sequence; \textsc{Accent} with levels \emph{ntr},
\emph{lc} and \emph{ec} coding which reading was cued by intonation;
\textsc{Position} with levels 1 to 4; and \textsc{Response} with
levels \emph{lc}, \emph{ec} and \emph{false}. The observed data under
this classification is given in Table~\ref{tab:TF-ContingencyTable}.
%
\begin{table}[h]

@@

\subfloat[][\lc neutral]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 & 30 &  6 & 1  \\
    Pos 2 & 24 &  9 & 4  \\
    Pos 3 & 25 &  8 & 3  \\
    Pos 4 & 11 & 10 & 1  \\ \addlinespace[0.15cm]
    Total & 90 & 33 & 9 
\end{tabular}}
%
\hspace{0.4cm}
%
\subfloat[][\ec neutral]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  28 &   4 &   5 \\ 
    Pos 2 &  26 &  10 &   1 \\ 
    Pos 3 &  19 &  11 &   6 \\ 
    Pos 4 &  11 &   9 &   2 \\ \addlinespace[0.15cm]
    Total &  65 &  34 &  14 
\end{tabular}}

\subfloat[][\lc with \lc cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  26 &   6 &   5 \\ 
    Pos 2 &  30 &   4 &   3 \\ 
    Pos 3 &  30 &   5 &   2 \\ 
    Pos 4 &  14 &   1 &   1 \\ 
    \addlinespace[0.15cm]
    Total & 100 & 16 & 11 
\end{tabular}}
%
\hspace{0.4cm}
%
\subfloat[][\ec with \lc cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  24 &   6 &   7 \\ 
    Pos 2 &  30 &   3 &   4 \\ 
    Pos 3 &  30 &   5 &   2 \\ 
    Pos 4 &  13 &   2 &   1 \\ 
    \addlinespace[0.15cm]
    Total &  97 &  16 &  14 
\end{tabular}}

\subfloat[][\lc with \ec cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  15 &  15 &   7 \\ 
    Pos 2 &  20 &  10 &   7 \\ 
    Pos 3 &  24 &   7 &   5 \\ 
    Pos 4 &   8 &   8 &   6 \\
    \addlinespace[0.15cm]
    Total &  67 & 40 &  25 
\end{tabular}}
%
\hspace{0.4cm}
%
\subfloat[][\ec with \ec cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  21 &  12 &   4 \\ 
    Pos 2 &  18 &  12 &   7 \\ 
    Pos 3 &  22 &  12 &   2 \\ 
    Pos 4 &  13 &   7 &   2 \\ 
    \addlinespace[0.15cm]
    Total &  74 &  43 &  15 
\end{tabular}}

@@

The best hierarchical log-linear model (by $AIC$) includes factors
\textsc{Response}, \textsc{Accent} and \textsc{Position}, as well as
the two-way interaction between \textsc{Response} and \textsc{Accent}
($\chi^2 = 48.802$, $df=60$, $p=0.85$, $AIC=72.802$ (compared to
$AIC=144$ of the standard model)). Crucially, the observed response
pattern does not differ significantly across conditions (i.e., which
reading can be judged first), but does depend on intonational cues.

The factor \textsc{Position} cannot be dropped without dramatically
reducing explanatory power of the model. However, inspecting the
coefficients of the above model, we see that the only role of factor
\textsc{Position} is to account for the relatively fewer data points
at position four. Moreover, if we abstract over intonation and
conditions, we get the following table:
\begin{center}
\centering
\begin{tabular}{rrrr}
  & \emph{lc} & \emph{ec} & \emph{false} \\ 
  \midrule
  Pos 1 & 144 & 49 & 29 \\ 
  Pos 2 & 148 & 48 & 26 \\ 
  Pos 3 & 150 & 48 & 20 \\ 
  Pos 4 &  70 & 37 & 13 \\ 
\end{tabular}
\end{center}
There is no interaction between \textsc{Position} and
\textsc{Response} ($\chi^2$, $df=6$, $p=0.3798$), suggesting that
\textsc{Position} does not affect the overal distribution of readings.

\todo[inline]{is this convincing to cast away \textsc{Position}
  effects? do we really need to care in this case? there is no
  particularly obvious difference across positions that one would need
  to worry about; also the way conditions were presented in the second
  block don't really lead to any obvious pattern either; in other
  words, neither does the faulty design suggest a positional pattern
  (or at least I can conceive of one), nor does there seem to be one
  (or at least I don't see one) }

\newpage

@@

If we only look at the judgement that subjects gave to the very first
target condition that they saw (be it \as or \es), we obtain the
following table:
\begin{center}
\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    accented & 15 &  5 & 0 &  9 \\
    neutral  &  6 &  0 & 0 &  2 \\ \addlinespace[0.15cm]
    Total    & 21 &  5 & 0 & 11
\end{tabular}
\end{center}
Notably, nobody gave a \emph{local} answer on the first encounter of a
target condition with neutral accentuation. Still, Pearson's $\chi^2$
test does not reject the null hypothesis of an equal distribution of
readings ($\chi^2 = 2.0546$, $p \approx 0.4318$).  \todo[inline]{Is
  this test applicable at all? Don't we have too little data points?
  How else could a dependence be tested in this case?}

>>>>

We excluded two participants from the analysis due to poor performance
on control conditions (VPs 2 and 29).

@@

\subfloat[][\as neutral]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 35 &  0 & 0 & 3 \\
    Pos 2 & 32 &  3 & 0 & 3 \\
    Pos 3 & 33 &  3 & 0 & 2 \\ \addlinespace[0.15cm]
    Total &100 &  6 & 0 & 8
\end{tabular}}
%
\hspace{0.2cm}
%
\subfloat[][\es neutral]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 19 &  5 & 0 & 14 \\
    Pos 2 & 25 &  2 & 0 & 11 \\
    Pos 3 & 21 &  6 & 0 & 11 \\ \addlinespace[0.15cm]
    Total & 65 & 13 & 0 & 36
\end{tabular}}

\subfloat[][\as accented]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 33 &  2 & 1 &  2 \\
    Pos 2 & 31 &  4 & 1 &  2 \\
    Pos 3 & 31 &  5 & 0 &  2 \\ \addlinespace[0.15cm]
    Total & 95 & 11 & 2 &  6
\end{tabular}}
%
\hspace{0.2cm}
%
\subfloat[][\es accented]{\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    Pos 1 & 24 &  8 & 0 &  6 \\
    Pos 2 & 22 &  7 & 0 &  9 \\
    Pos 3 & 27 &  6 & 0 &  5 \\ \addlinespace[0.15cm]
    Total & 73 & 21 & 0 & 20
\end{tabular}}

@@

To test which factors are relevant to explain the data distribution in
Table~\ref{tab:CC-ContingencyTable}, we compute log-linear models
\citep{KnokeBurke1980:Log-Linear-Mode}. The relevant factors are
\textsc{Position}, \textsc{Response}, \textsc{Condition}, and
\textsc{Accent}. As there are hardly any \emph{global} responses, we
must lump levels \emph{global} and \emph{false} of factor
\textsc{Response} into a single level \emph{other}, because otherwise
log-linear regression is not applicable. We select the ``best''
predictor of the data by a gradient search over hierarchical models in
terms of AICs. The best model takes factors \textsc{Accent},
\textsc{Response} and \textsc{Condition} into account, as well as the
interaction terms \textsc{Response:Accent} and
\textsc{Response:Condition} ($\chi^2 = 14.29$, $df=27$, $p = 0.98$,
$AIC = 32.29$ (compared to $AIC=72$ of the saturated
model)). Crucially, factor \textsc{Position} is spurious. Response
patterns only depended on the type of the sentence and whether it was
produced with accentuation.

\todo[inline]{ what's the argument to accept that there are local readings
    for neutral accentuation in the first place? ; make plots with ``raw responses'' on position
}

@@

For our preference-related control conditions the relevant factors
are: \textsc{Condition} with levels \emph{lc} and \emph{ec} coding
which reading could be judged first in a sequence, \textsc{Accent}
with levels \emph{ntr}, \emph{lc} and \emph{ec} coding which reading
was cued by intonation, \textsc{Response} with levels \emph{lc},
\emph{ec} and \emph{false}, and finally \textsc{Position} encoding how
frequently subjects had seen a given sentence-accentuation pair
before. Some subjects saw three, others four trials, so that to ensure
unbiased applicability of our log-linear models, we excluded every
fourth encounter of the same sentence type and accentuation. The
observed data under this classification is given in
Table~\ref{tab:TF-ContingencyTable}.
%
\begin{table}[h]

@@

\subfloat[][\lc neutral]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 & 23 &  9 & 6  \\
    Pos 2 & 23 & 11 & 4  \\
    Pos 3 & 27 &  8 & 3  \\ \addlinespace[0.15cm]
    Total & 73 & 28 & 13 
\end{tabular}}
%
\hspace{0.4cm}
%
\subfloat[][\ec neutral]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  22 &   5 &  11 \\ 
    Pos 2 &  28 &   5 &   5 \\ 
    Pos 3 &  29 &   7 &   2 \\ \addlinespace[0.15cm]
    Total &  79 &  17 &  18 
\end{tabular}}

\subfloat[][\lc with \lc cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  26 &   6 &   6 \\ 
    Pos 2 &  30 &   5 &   3 \\ 
    Pos 3 &  34 &   2 &   2 \\ 
    \addlinespace[0.15cm]
    Total & 90  & 13 & 11 
\end{tabular}}
%
\hspace{0.4cm}
%
\subfloat[][\ec with \lc cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  30 &   4 &   4 \\ 
    Pos 2 &  33 &   4 &   1 \\ 
    Pos 3 &  31 &   4 &   3 \\ 
    \addlinespace[0.15cm]
    Total &  94 &  12 &  8 
\end{tabular}}

\subfloat[][\lc with \ec cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  15 &  18 &   5 \\ 
    Pos 2 &  16 &  16 &   6 \\ 
    Pos 3 &  15 &  16 &   6 \\
    \addlinespace[0.15cm]
    Total &  46 & 50 &  17 
\end{tabular}}
%
\hspace{0.4cm}
%
\subfloat[][\ec with \ec cue]{\begin{tabular}{lccc}
    & \emph{lc} & \emph{ec} & \emph{false} \\
    \midrule
    Pos 1 &  18 &  12 &   8 \\ 
    Pos 2 &  18 &   9 &  11 \\ 
    Pos 3 &  18 &  10 &  10 \\
    \addlinespace[0.15cm]
    Total &  54 &  31 &  29 
\end{tabular}}

@@

Just like for the target conditions, the best hierarchical log-linear
model (by $AIC$) includes factors \textsc{Response}, \textsc{Accent}
and \textsc{Condition}, as well as the two-way interactions
\textsc{Response:Accent} and \textsc{Response:Condition} ($\chi^2 =
23.14$, $df=42$, $p=0.99$, $AIC=47.14$ (compared to $AIC=108$ of the
saturated model)). Crucially, the observed response pattern does not
differ significantly across conditions (i.e., which reading can be
judged first), but does depend on intonational cues.

@@

If we only look at the judgement that subjects gave to the very first
target condition that they saw (be it \as or \es), we obtain the
following table:
\begin{center}
\begin{tabular}{lcccc}
    & \emph{literal} & \emph{local} & \emph{global} & \emph{false} \\
    \midrule
    accented & 10 &  3 & 1 &  2 \\
    neutral  & 15 &  0 & 0 &  7 \\ \addlinespace[0.15cm]
    Total    & 25 &  3 & 0 &  9
\end{tabular}
\end{center}
Notably, nobody gave a \emph{local} answer on the first encounter of a
target condition with neutral accentuation. Pearson's $\chi^2$ test
does reject the null hypothesis of an equal distribution of readings
($\chi^2 = 7.0051$, $p \approx 0.034$).
